---
title: "Estimate the cell type composition"
format: revealjs
editor: visual
---

## Research question

Are the methylation levels confounded by the cell type composition?

## Estimate cell counts
```{r}
library(FlowSorted.Blood.450k)
pData(age.rgSet)$Slide <- as.numeric(pData(age.rgSet)$Slide) # Make sure the "Slide" column in the rgSet phenotype data contains numerical values to prevent encountering an error when estimating cell counts with estimateCellCounts()

cell.counts <- estimateCellCounts(age.rgSet)
```

## Visualize cell type proportions by age

If we're dealing with samples obtained from a population of mixed cells, like blood, it's a good idea to examine the possibility of any relationship between variations in the proportions of different cell types and the variable you're studying (e.g., age group).

```{r}
par(mfrow=c(1,1))
age.pal <- brewer.pal(8,"Set1")

counts.newborn = cell.counts[age.sample.info$Sample_Group=="NewBorns",]
counts.old = cell.counts[age.sample.info$Sample_Group=="OLD",]
boxplot(counts.newborn, at=0:5*3 + 1, xlim=c(0, 18), 
        ylim=range(counts.newborn, counts.old), xaxt="n", 
        col=age.pal[1], main="", ylab="Cell type proportion")
# at - numeric vector giving the locations where the boxplots should be drawn, particularly when add = TRUE; defaults to 1:n where n is the number of boxes.
# xaxt - A character which specifies the x axis type. Specifying "n" suppresses plotting of the axis.
boxplot(counts.old, at=0:5*3 + 2, xaxt="n", add=T, col=age.pal[2])
axis(1, at=0:5*3 + 1.5, labels=colnames(counts.newborn), tick=T)
# colnames(counts.newborn) # [1] "CD8T"  "CD4T"  "NK"    "Bcell" "Mono"  "Gran" 
legend("topleft", legend=c("NewBorns","OLD"), fill=age.pal)
```

The above plot illustrates a significant association between variations in blood cell type proportions and age within this methylation dataset. Conducting cell composition estimation can be helpful to spot the potential confounding issues when analyzing a dataset comprising various cell types.

