---
title: "ATAC-Seq Data Analysis with Human Data - Align the FASTQ files"
author: "Anni Liu"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: show
---

```{r, shorcut, include=FALSE}
## RStudio keyboard shortcut
# Cursor at the beginning of a command line: Ctrl+A
# Cursor at the end of a command line: Ctrl+E
# Clear all the code from your console: Ctrl+L
# Create a pipe operator %>%: Ctrl+Shift+M (Windows) or Cmd+Shift+M (Mac)
# Create an assignment operator <-: Alt+- (Windows) or Option+- (Mac) 
# Knit a document (knitr): Ctrl+Shift+K (Windows) or Cmd+Shift+K (Mac)
# Comment or uncomment current selection: Ctrl+Shift+C (Windows) or Cmd+Shift+C (Mac)
```

# Run the FastQC to check the read quality in FASTQ files
```{bash}
zcat *.fastq.gz | fastqc stdin --outdir=../report/
# zcat ATACSample_r1.fastq.gz ATACSample_r2.fastq.gz | fastqc stdin --outdir=../report/
```

# Align the FASTQ files
## Create a reference genome
```{r}
library(BSgenome.Hsapiens.UCSC.hg19)
mainChromosomes <- paste0("chr", c(1:21,"X","Y","M"))
mainChrSeq <- lapply(mainChromosomes,
                     function(x) BSgenome.Hsapiens.UCSC.hg19[[x]])
names(mainChrSeq) <- mainChromosomes
mainChrSeqSet <- DNAStringSet(mainChrSeq)
Biostrings::writeXStringSet(x = mainChrSeqSet, 
                            filepath = "BSgenome.Hsapiens.UCSC.hg19.mainChrs.fa")
```

## Update the reference genome based on hg38
```{r}
# https://gatk.broadinstitute.org/hc/en-us/articles/360035890951-Human-genome-reference-builds-GRCh38-or-hg38-b37-hg19
library(BSgenome.Hsapiens.UCSC.hg38)
main_chromosomes <- paste0("chr", c(1:22, "X", "Y", "M"))
main_chr_seq <- lapply(main_chromosomes,
                       function(x) BSgenome.Hsapiens.UCSC.hg38[[x]])
names(main_chr_seq) <- main_chromosomes
main_chr_seq_set <- DNAStringSet(main_chr_seq)
Biostrings::writeXStringSet(x = main_chr_seq_set, 
                            filepath = "BSgenome.Hsapiens.UCSC.hg38.mainchr.fa")
```

## Update the reference genome based on hs1
```{r}
library(BSgenome.Hsapiens.UCSC.hs1)
main_chromosomes <- paste0("chr", c(1:22, "X", "Y", "M"))
main_chr_seq <- lapply(main_chromosomes,
                       function(x) BSgenome.Hsapiens.UCSC.hs1[[x]])
names(main_chr_seq) <- main_chromosomes
main_chr_seq_set <- DNAStringSet(main_chr_seq)
Biostrings::writeXStringSet(x = main_chr_seq_set, 
                            filepath = "BSgenome.Hsapiens.UCSC.hs1.mainchr.fa")
```


## Build the index
```{r}
library(Rsubread)
buildindex(basename = "BSgenome.Hsapiens.UCSC.hg19.mainChrs",
           reference = "BSgenome.Hsapiens.UCSC.hg19.mainChrs.fa",
           indexSplit = TRUE,
           memory = 1000)

# indexSplit: logical indicating whether the index can be split into multiple blocks. The block size is determined by the value of memory. FALSE by default (ie. a single-block index is generated).
# memory: a numeric value specifying the amount of memory (in megabytes) used for storing the index during read mapping. 8000 MB by default. Note that this option is ignored when indexSplit is FALSE.
```


## Align **paired-end** sequence reads from Greenleaf
```{r}
# Take a peek at 2 fastq files
library(ShortRead)
# The following 2 fastq files data downloaded from https://www.ebi.ac.uk/ena/browser/view/SRR891269
read1 <- readFastq(dirPath = "./data/ATACSample_r1.fastq.gz")
read2 <- readFastq(dirPath = "./data/ATACSample_r2.fastq.gz")
id(read1)[1]
# BStringSet object of length 1:
#     width seq
# [1]    59 HISEQ:236:HA03EADXX:1:1101:1147:2237 1:Y:0:TAAGGCGACTCTCTAT
id(read2)[1]
# BStringSet object of length 1:
#     width seq
# [1]    59 HISEQ:236:HA03EADXX:1:1101:1147:2237 2:Y:0:TAAGGCGACTCTCTAT
# These 2 IDs indicate 2 reads are paired, stemming from the same fragment

library(Rsubread)
align(index = "BSgenome.Hsapiens.UCSC.hg19.mainChrs",
      readfile1 = "./data/ATACSample_r1.fastq.gz", 
      readfile2 = "./data/ATACSample_r2.fastq.gz",
      output_file = "ATAC_50K_2.bam",
      nthreads = 2, type = 1,
      unique = TRUE, maxFragLength = 2000)
# type: a character string or an integer giving the type of sequencing data. Possible values include rna (or 0; RNA-seq data) and dna (or *1; genomic DNA-seq data such as WGS, WES, *ChIP-seq data etc.). Character strings are case insensitive.
# unique: logical indicating if only uniquely mapped reads should be reported. A uniquely mapped read has one single mapping location that has less mis-matched bases than any other candidate locations. By default, multi-mapping reads will be reported in addition to uniquely mapped reads. Number of alignments reported for each multi-mapping read is determined by the nBestLocations parameter.
# maxFragLength: numeric value giving the maximum fragment length. 600 by default.
```

## Perform the alignment relative to hg38 using the new ATACseq data from female lung
```{bash}
# https://www.ebi.ac.uk/ena/browser/view/SAMN08743398?show=reads
wget -nc ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR687/008/SRR6870408/SRR6870408_2.fastq.gz
wget -nc ftp://ftp.sra.ebi.ac.uk/vol1/fastq/SRR687/008/SRR6870408/SRR6870408_1.fastq.gz
# -nc: if a file with the same name already exists, wget will not overwrite it but will skip the download.
```

```{r}
r1 <- readFastq(dirPath = "./data/SRR6870408_1.fastq.gz")
r2 <- readFastq(dirPath = "./data/SRR6870408_2.fastq.gz")
id(r1)[1]
# BStringSet object of length 1:
#     width seq
# [1]    16 SRR6870408.1 1/1
id(r2)[1]
# BStringSet object of length 1:
#     width seq
# [1]    16 SRR6870408.1 1/2

align(index = "BSgenome.Hsapiens.UCSC.hg38.mainchr",
      readfile1 = "./data/SRR6870408_1.fastq.gz", 
      readfile2 = "./data/SRR6870408_2.fastq.gz",
      output_file = "ATAC_female_lung.bam",
      nthreads = 4, type = 1,
      unique = TRUE, maxFragLength = 600)
```


## Alternative alignment approach: `Rbowtie2`
### Build the Rbowtie2 index
```{r}
library(Rbowtie2)
bowtie2_build(references = "BSgenome.Hsapiens.UCSC.hg19.mainChrs.fa",
              bt2Index = "BSgenome.Hsapiens.UCSC.hg19.mainChrs_bowtie2")
```

```{r}
bowtie2_build(references = "BSgenome.Hsapiens.UCSC.hg38.mainchr.fa",
              bt2Index = "BSgenome.Hsapiens.UCSC.hg38.mainchr_bowtie2")
```

### Decompress the FASTQ files
```{bash}
gunzip ./ATAC_Data/ATAC_FQs/SRR891269_1.fastq.gz
gunzip ./ATAC_Data/ATAC_FQs/SRR891269_2.fastq.gz

# Keep the compressed file: 
# gunzip -c ./ATAC_Data/ATAC_FQs/SRR891269_1.fastq.gz > ./ATAC_Data/ATAC_FQs/SRR891269_1.fastq
```

```{bash}
gunzip ./data/SRR6870408_1.fastq.gz
gunzip ./data/SRR6870408_2.fastq.gz
```

### Align the paired reads
```{r}
bowtie2(bt2Index = "BSgenome.Hsapiens.UCSC.hg19.mainChrs_bowtie2",
        samOutput = "ATAC_50K_2_bowtie2.sam",
        seq1 = "./ATAC_Data/ATAC_FQs/SRR891269_1.fastq",
        seq2 = "./ATAC_Data/ATAC_FQs/SRR891269_2.fastq")

# Convert the output SAM file to a more useable BAM file
library(Rsamtools)
asBam("ATAC_50K_2_bowtie2.sam") # asBam, asSam return the file name of the destination file.

# To save the disk space, we are removing the uncompressed FASTQ and SAM files, while retaining the BAM file
unlink(c("./ATAC_Data/ATAC_FQs/SRR891269_1.fastq",
         "./ATAC_Data/ATAC_FQs/SRR891269_2.fastq",
         "ATAC_50K_2_bowtie2.sam"))
```

```{r}
bowtie2(bt2Index = "BSgenome.Hsapiens.UCSC.hg38.mainchr_bowtie2",
        samOutput = "./data/ATAC_female_lung_bowtie2.sam",
        seq1 = "./data/SRR6870408_1.fastq",
        seq2 = "./data/SRR6870408_2.fastq")
asBam("./data/ATAC_female_lung_bowtie2.sam") 

unlink(c("./data/SRR6870408_1.fastq",
         "./data/SRR6870408_2.fastq",
         "./data/ATAC_female_lung_bowtie2.sam"))
```


## Sort and index
```{r}
library(Rsamtools)
sortedBAM <- file.path(dirname(outBAM),
                       paste0("Sorted_", basename(outBAM))) # Construct the path to a file from components 

sortBam(outBAM, gsub("\\.bam", "", basename(sortedBAM)))
indexBam(sortedBAM)
```


## Check the distribution of mapped reads across chromosomes
```{r}
library(Rsamtools)
mappedReads <- idxstatsBam(sortedBAM)
save(mappedReads, file = "data/pres1_idxstats.RData")

load("data/pres1_idxstats.RData")
```

```{r data.in.package}
file.path <- system.file("extdata", "ex1.bam", package = "Rsamtools", mustWork = T)
indexBam(file.path)
idxstatsBam(file.path)
# seqnames: chromosome names; mapped: how many reads are mapped on every chromosome
```

```{r visual}
library(ggplot2)
load("data/pres1_idxstats.RData")
ggplot(data = mappedReads, 
       mapping = aes(x = seqnames, y = mapped, fill = seqnames))+
  geom_bar(stat = "identity") + 
  geom_text(mapping = aes(label = mapped), vjust = 0, color = "black", size = 3.5)+
  coord_flip()
# There are lots of mitochondrial reads
```
